Making Sequential Programs Deterministic

### 2019-04-03 design.tex
Edits in section 2.4

%% This section is too basic, and therefore is boring because most readers
%% will already know it.  It also delays the interesting parts of the paper
%% for too long.  We can introduce concepts as they are needed in the rest
%% of the paper.
% \subsection{Type checking background}\label{type-checking}
% The major components of a type system include: 1) \textit{types}, 2) \textit{subtyping rules}, and 3) \textit{dataflow analysis}.
% A \textit{type} serves as an abstraction for the set of acceptable values for any expression. The types in a type system form a lattice of finite height. 
% The hierarchy of types in this lattice defines subtyping relationships among them.
% In our framework, we require every type hierarchy to define a unique\<@Top> and a \<@Bottom> element. This ensures that
% any given pair of types has a \textit{least upper bound} and a \textit{greatest lower bound}.
% Throughout this paper, we use the notation \<@A> <: \<@B> to denote that type \<@A> is a subtype of type \<@B>.
% As an illustration of the type hierarchy, consider the lattice of types shown in \cref{fig-example-lattice}.
% \begin{figure}
%     \begin{center}
%         \includegraphics[scale=0.15]{lattice}
%     \end{center}
%     \caption{Example type hierarchy}
%     \label{fig-example-lattice}
% \end{figure}
% In this type system, \<@C> is a subtype of \<@A> as well as \<@B> and types \<@A> and \<@B> are incomparable.
% The type checker now performs additional type checking (similar to javac's type checking) with respect to this type hierarchy
% and reports any violations. The snippet in \cref{code-invalid1} would result in an error at the assignment statements:
% \begin{figure}
%     \begin{verbatim}
%     @A int x; @B int y; @C int z;
%     x = y;    // Because types @A and @B are incomparable.
%     z = x;    // Because @C is a subtype of @A.
%     \end{verbatim}
% \caption{Example: Invalid assingment}
% \label{code-invalid1}
% \end{figure}
% 
% %Method overriding and example
% For method overriding, the usual rules apply for parameters and arguments i.e covariant subtyping for parameters
% and contravariant subtyping for return types. Consider the example in \cref{code-invalid2} where method \<foo> of \<class X> is overridden
% in \<class Y>.
% \begin{figure}
%     \begin{verbatim}
%     class X {
%         @B int foo(@C int param) {}
%     }
%     class Y extends X {
%         @C int foo(@B int param) {}
%     }
%     \end{verbatim}
%     \caption{Example: Invalid Method override}
%     \label{code-invalid2}
% \end{figure}
% 
% The overriding method is invalid for two reasons: type of the parameter (\<@B>) is not a subtype of the parameter type in the overridden method (\<@C>) and the return type (\<@C>) is not a supertype of the return type of the overridden method (\<@B>).
% 
% %collections type parameters - invariant
% Collection types are invariant. So, if a \<List> x is declared as \codeid{@A List<@A String> x;} and another \<List y> is declared
% as  \codeid{@A List<@B String> y;}, writing \<x = y> would result in an error.
% 
% %covariant array types
% Arrays follow covariant subtyping rules in java. Therefore, the assignment statement \<@A int @A[] x = y;> where \<y> was 
% declared as \<@B int @A[] y> would type check without any warning.

@@

\Cref{type-validity} shows the type well formedness
%\todo{The caption says
%  ``type well formedness''.  Choose one term and use it consistently
%  throughout.}
rules of \ourTypeSystem. 
As stated in the rule \rulename{noncollection}, the types \<Det> and \<NonDet>
can be written on any non-collection Java basetype.

@@

\subsubsection{Behavior of order-nondeterministic collections}

@@

To state the first point more formally, the 
typical type for a list access operation, such as \<get>, is
$$\forall \tau . \<List>\angles{\tau} \times \<Det Int> \rightarrow \tau$$
but this type is incorrect for \ourTypeSystem.  Each of the following is a
correct (but overly restrictive) type for \<get>:
\begin{eqnarray*}
\forall \kappa\ \beta . \<NonDet>\ \<List>\angles{\kappa\ \beta} \times \<Det Int>
& \strut\hspace{-1em} \rightarrow \<NonDet>\ \beta
& \mbox{\quad in this case $\kappa = \<NonDet>$} \\
\forall \kappa\ \beta . \<OrderNonDet>\ \<List>\angles{\kappa\ \beta} \times \<Det Int>
& \strut\hspace{-1em} \rightarrow \<NonDet>\ \beta
& \mbox{\quad in this case $\kappa \in \{ \<OrderNonDet>, \<Det> \} $} \\
\forall \kappa\ \beta . \<Det>\ \<List>\angles{\kappa\ \beta} \times \<Det Int>
& \strut\hspace{-1em} \rightarrow \<Det>\ \beta\phantom{\<Non>}
& \mbox{\quad in this case $\kappa = \<Det>$}
\end{eqnarray*}

@@

    \rulename{iteration}: 
    $
    \forall \tau . \|OrderNonDet \ C| \angles{\tau} \rightarrow 
    \|OrderNonDet \ Iterator| \angles{\tau}
    $

@@

    \rulename{next element}:
    $
    \forall \kappa \beta . \|OrderNonDet \ Iterator|\angles{\kappa \ \beta} \rightarrow
    \|NonDet|\ \beta
    $

@@

    \rulename{hasnext element}:
    $
    \forall \tau . \|OrderNonDet \ Iterator|\angles{\tau} \rightarrow
    \|Det \ boolean|
    $
    \bigskip

@@

    \rulename{size}:
    $
    \forall \tau . \|OrderNonDet \ C|\angles{\tau} \rightarrow
    \|Det \ int|
    $

@@

\todo{Cut parts of the below; show the rest in Greek.}

@@

\todo{Greekify the following}
\begin{itemize}
\item
Type polymorphism has the usual semantics.  For example, the type of the
identity function is $\forall \tau . \tau \rightarrow \tau$, which can be
equivalently written as
$\forall \kappa \ \beta . \kappa \ \beta \rightarrow \kappa \ \beta$.
\item
Basetype polymorphism lets the basetype vary without applies a

An example of a type-polymorphic method is identity

In \theDeterminismCheckerImplementation,
type polymorphism is handled by Java's generics mechanism, which
\theDeterminismChecker fully supports, including class and method generics,
inference, etc.
Given the Java declaration
\begin{Verbatim}
<T> T identity(T arg) { return arg; }
\end{Verbatim}
the type of \<identity(arg)> is the same as the type of
\<arg>, and a programmer can write a type such as \codeid{List<@NonDet
  Integer>} (but see below for details about collections).
\item
Basetype polymorphism is enabled by writing a type qualifier on a use of a
type variable, which overrides the type qualifier at the instantiation
site.
For example, a \<choose> operation on sets could be defined in Java as
\begin{Verbatim}
class ChooseableSet<T> implements Set<T> {
  // return an arbitrary element from the set
  @NonDet T choose();
}
\end{Verbatim}
\item
Java does not provide a syntax that can be used for qualifier polymorphism,
so we use a special type qualifier name, \<@PolyDet>.
A qualifier-polymorphic method \<m> with signature $\forall \kappa . \kappa\ \<int> \times \<Det boolean> \rightarrow
\kappa\ \<String>$ would be written in our dialect of Java as
\begin{Verbatim}
@PolyDet String m(@PolyDet int, @Det boolean)
\end{Verbatim}
Each use of \<@PolyDet> stands for a use of the qualifier variable
$\kappa$, and there is no need to declare the qualifier variable $\kappa$.
\TheDeterminismChecker currently supports
qualifier
polymorphism on methods but not on classes.
Therefore, \<@PolyDet> may be written in methods (signatures and bodies)
but not on fields.
\end{itemize}

When a user writes a polymorphic qualifier on a method signature or a type parameter,
it indicates that it could be instantiated
%\todo{Should all occurrences of ``resolve'' be ``can be instantiated''} 
to any type in the type system depending on how it is used.
In \theDeterminismChecker, we define a polymorphic qualifier \<PolyDet>.
One of the most common locations for a polymorphic qualifier is at a method signature.
Consider the following declaration of addition:
\begin{verbatim}
@PolyDet int plus(@PolyDet int a, @PolyDet int b) { return a+b; }
\end{verbatim}
\todo{It has the standard meaning.  There are two ways to interpret this.}
There are two valid instantiations for this declaration:
\begin{Verbatim}
  @NonDet int plus(@NonDet int a, @NonDet int b) { return a+b; }
  @Det int plus(@Det int a, @Det int b) { return a+b; }
\end{Verbatim}
This indicates that method \<foo> can be called with arguments having any of the \codeid{Det or NonDet} type qualifiers.
%   (\<@OrderNonDet> is not allowed because it is invalid on primitive types).
 \<PolyDet> gets instantiated to the least upper bound of
the actual types on arguments. For instance, if this method is called as \codeid{foo(@Det String arg1, @NonDet boolean arg2)}, \theDeterminismChecker instantiates the method declaration as \codeid{@NonDet int foo(@NonDet String param1, @NonDet boolean param2)}
causing the return type to have the type qualifier \<NonDet>.

\subsection{Polymorphism rules for collections}\label{polymorphism-up-down}

Some types are invalid and must not be permitted (during polymorphic instantiation).

Recall that \theDeterminismChecker does not permit \<OrderNonDet> on a type
that isn't a collection.
As a consequence, resolution of polymorphic type qualifiers at method signatures could have undesirable effects.
Suppose a method declaration is annotated as \codeid{PolyDet int size(PolyDet List<T> list)}. Invoking this method
with an argument of type \codeid{OrderNonDet List<Det String>} will result in \<PolyDet> getting instantiated to \<OrderNonDet> which is invalid
on the return type \<int>. We circumvent this scenario by introducing two variants to the \<PolyDet> qualifier, \PolyDetUp
and \PolyDetDown.

@@

The semantics of these variants are as follows:
\begin{itemize}
    \item \PolyDetUp: Same as \<PolyDet> except when it is instantiated as \<OrderNonDet>, it is replaced with \<NonDet>.
    \item \PolyDetDown: Same as \<PolyDet> except when it is instantiated as \<OrderNonDet>, it is replaced with \<Det>.
\end{itemize} 
The modifiers $\uparrow$ and $\downarrow$ are indicative of the subtyping relationship of \<OrderNonDet> with \<NonDet> and \<Det> respectively
in the determinism type hierarchy.

%\Cref{fig-poly-resolution} presents the rule for type checking of methods annotated with \<PolyDet> or its variants.
A polymorphic method type checks if every instantiation of \<PolyDet> type checks. If this method has \PolyDetUp\ or \PolyDetDown\
in its signature then the instantiation of \<PolyDet> at that location gets replaced by the appropriate type as shown in \cref{fig:poly-resolutions}.

Annotating \<size()> as \codeid{PolyDet$\downarrow$ int size(PolyDet List<T> list)} will only allow the following type instantiations:
\begin{itemize}
    \item \codeid{@Det int size(@Det List<T> list)}
    \item \codeid{@Det int size(@OrderNonDet List<T> list)}
    \item \codeid{@NonDet int size(@NonDet List<T> list)}
\end{itemize}

%\begin{figure}
%    $\infer[\rulename{poly resolution}]{\pi_r \ \beta_r\ \<m>(\overline{\pi_i \ \beta_i})\{\ \}}{\forall \kappa \in \{\|NonDet, OrderNonDet, Det|\} : \kappa \ \beta_r\ \<m>(\overline{K \ \beta_i})\{\ \} & \pi_r, \pi_i \in \{\|PolyDet|, \PolyDetUp, \PolyDetDown    \}}$
%    \caption{Polymorphic resolution rule}
%    \todo{I'm not sure what this is doing.  What code does it reject?}
%    \todo{This suggests that \PolyDetUp is a qualifier, rather than
%      $\uparrow$ being an operator on qualifiers.  We should clarify which
%      it is.}
%    \label{fig-poly-resolution}
%\end{figure}

@@

\begin{figure}
    $\infer[\rulename{ordernondet polyup}]
    {\tau \uparrow \ = \|NonDet \ \beta|}
    {\tau \ = \|OrderNonDet \ \beta|
        & \wellformed{\tau}
    }$

@@

    $\infer[\rulename{polyup}]
    {\tau \uparrow \ = \kappa \ \beta}
    {\tau \ =\kappa \ \beta
        & \kappa \in \{\|Det, NonDet|\}
        & \wellformed{\tau}
    }$

@@

    $\infer[\rulename{ordernondet polydown}]
    {\tau \downarrow \ = \|Det \ \beta|}
    {\tau \ = \|OrderNonDet \ \beta|
        & \wellformed{\tau}
    }$

@@

    $\infer[\rulename{polydown}]
    {\tau \downarrow \ = \kappa \ \beta}
    {\tau \ =\kappa \ \beta
        & \kappa \in \{\|Det, NonDet|\}
        & \wellformed{\tau}
    }$
\caption{Polymorphic resolution ($\uparrow$ is overloaded for types and qualifiers)}
\label{fig:poly-resolutions}
\end{figure}

@@

In order to provide soundness guarantees, we must ensure that a collection that is
declared as \<Det> is not side-effected in non-deterministic ways. Consider the following qualifier 
on the List add method: \codeid{public void add(PolyDet ArrayList<E> this, PolyDet int index, E element)}.
Suppose a client calls this method as follows:
\begin{verbatim}
@Det ArrayList<@Det String> list;
@NonDet int random;
@Det String str;
list.add(random, str);
\end{verbatim}
This will result in the instantiation \codeid{public void add(@NonDet ArrayList<E> this, @NonDet int index, E element)} and
the method invocation will type check. This is problematic because even though \<list> was declared to be \<Det>,
the type checker allows the addition of an element at a \<NonDet> location violating our determinism guarantees.
We prevent this behavior by introducing another variant of \<PolyDet>, \<PolyDet("use")> that does not affect
type instantiations. This is especially useful in preventing methods from non-deterministically modifying the state
of a deterministic receiver. To have this effect, the library method \<ArrayList.add> is annotated as:
\codeid{public void add(PolyDet ArrayList<E> this, PolyDet("use") int index, E element)}.
Since \<PolyDet("use")> has no effect on type instantiations, calling this method on a \<Det List>
will result in the method declaration being instantiated as \codeid{public void add(@Det ArrayList<E> this, @Det int index, E element)}
irrespective of the type of \<index>, thereby preventing undesired side-effects.

>>>>

%% This section is too basic, and therefore is boring because most readers
%% will already know it.  It also delays the interesting parts of the paper
%% for too long.  We can introduce concepts as they are needed in the rest
%% of the paper.
% \subsection{Type checking background}\label{type-checking}
% The major components of a type system include: 1) \textit{types}, 2) \textit{subtyping rules}, and 3) \textit{dataflow analysis}.
% A \textit{type} serves as an abstraction for the set of acceptable values for any expression. The types in a type system form a lattice of finite height. 
% The hierarchy of types in this lattice defines subtyping relationships among them.
% In our framework, we require every type hierarchy to define a unique\<@Top> and a \<@Bottom> element. This ensures that
% any given pair of types has a \textit{least upper bound} and a \textit{greatest lower bound}.
% Throughout this paper, we use the notation \<@A> <: \<@B> to denote that type \<@A> is a subtype of type \<@B>.
% As an illustration of the type hierarchy, consider the lattice of types shown in \cref{fig-example-lattice}.
% \begin{figure}
%     \begin{center}
%         \includegraphics[scale=0.15]{lattice}
%     \end{center}
%     \caption{Example type hierarchy}
%     \label{fig-example-lattice}
% \end{figure}
% In this type system, \<@C> is a subtype of \<@A> as well as \<@B> and types \<@A> and \<@B> are incomparable.
% The type checker now performs additional type checking (similar to javac's type checking) with respect to this type hierarchy
% and reports any violations. The snippet in \cref{code-invalid1} would result in an error at the assignment statements:
% \begin{figure}
%     \begin{verbatim}
%     @A int x; @B int y; @C int z;
%     x = y;    // Because types @A and @B are incomparable.
%     z = x;    // Because @C is a subtype of @A.
%     \end{verbatim}
% \caption{Example: Invalid assignment}
% \label{code-invalid1}
% \end{figure}
% 
% %Method overriding and example
% For method overriding, the usual rules apply for parameters and arguments i.e covariant subtyping for parameters
% and contravariant subtyping for return types. Consider the example in \cref{code-invalid2} where method \<foo> of \<class X> is overridden
% in \<class Y>.
% \begin{figure}
%     \begin{verbatim}
%     class X {
%         @B int foo(@C int param) {}
%     }
%     class Y extends X {
%         @C int foo(@B int param) {}
%     }
%     \end{verbatim}
%     \caption{Example: Invalid Method override}
%     \label{code-invalid2}
% \end{figure}
% 
% The overriding method is invalid for two reasons: type of the parameter (\<@B>) is not a subtype of the parameter type in the overridden method (\<@C>) and the return type (\<@C>) is not a supertype of the return type of the overridden method (\<@B>).
% 
% %collections type parameters - invariant
% Collection types are invariant. So, if a \<List> x is declared as \codeid{@A List<@A String> x;} and another \<List y> is declared
% as  \codeid{@A List<@B String> y;}, writing \<x = y> would result in an error.
% 
% %covariant array types
% Arrays follow covariant subtyping rules in java. Therefore, the assignment statement \<@A int @A[] x = y;> where \<y> was 
% declared as \<@B int @A[] y> would type check without any warning.

@@

\Cref{type-validity} shows the type well formedness
%\todo{The caption says
%  ``type well formedness''.  Choose one term and use it consistently
%  throughout.}
rules of \ourTypeSystem. 
(For brevity, we use ``valid'' synonymously with ``well-formed'' and
``invalid'' synonymously with ``non-well-formed''.)
As stated in the rule \rulename{noncollection}, the types \<Det> and \<NonDet>
can be written on any non-collection Java basetype.

@@

\subsubsection{Behavior of order-nondeterministic collections}\label{sec:ond-behavior}

@@

To state the first point more formally, the 
typical type for a list access operation, such as \<get>, is
$$\forallt{\tau} \<List>\angles{\tau} \times \<Det Int> \rightarrow \tau$$
but this type is incorrect for \ourTypeSystem.  Each of the following is a
correct (but overly restrictive) type for \<get>:
\begin{eqnarray*}
\forallt{\kappa\ \beta} \<NonDet>\ \<List>\angles{\kappa\ \beta} \times \<Det Int>
& \strut\hspace{-1em} \rightarrow \<NonDet>\ \beta
& \mbox{\quad in this case $\kappa = \<NonDet>$} \\
\forallt{\kappa\ \beta} \<OrderNonDet>\ \<List>\angles{\kappa\ \beta} \times \<Det Int>
& \strut\hspace{-1em} \rightarrow \<NonDet>\ \beta
& \mbox{\quad in this case $\kappa \in \{ \<OrderNonDet>, \<Det> \} $} \\
\forallt{\kappa\ \beta} \<Det>\ \<List>\angles{\kappa\ \beta} \times \<Det Int>
& \strut\hspace{-1em} \rightarrow \<Det>\ \beta\phantom{\<Non>}
& \mbox{\quad in this case $\kappa = \<Det>$}
\end{eqnarray*}

@@

    \rulename{iteration}: 
    $
    \forallt{\tau} \|OrderNonDet \ C| \angles{\tau} \rightarrow 
    \|OrderNonDet \ Iterator| \angles{\tau}
    $

@@

    \rulename{next element}:
    $
    \forallt{kappa \beta} \|OrderNonDet \ Iterator|\angles{\kappa \ \beta} \rightarrow
    \|NonDet|\ \beta
    $

@@

    \rulename{hasnext element}:
    $
    \forallt{\tau} \|OrderNonDet \ Iterator|\angles{\tau} \rightarrow
    \|Det \ boolean|
    $
    \bigskip

@@

    \rulename{size}:
    $
    \forallt{\tau} \|OrderNonDet \ C|\angles{\tau} \rightarrow
    \|Det \ int|
    $

@@

\todo{Can we come up with more examples of interesting functions to
  describe here?  Or put them in section 3.}

@@

\todo{Greekify the following}
\begin{itemize}
\item
Type polymorphism has the usual semantics.  For example, the type of the
identity function is $\forallt{\tau} \tau \rightarrow \tau$, which can be
equivalently written as
$\forallt{\kappa \ \beta} \kappa \ \beta \rightarrow \kappa \ \beta$.
(Compare this to 
$\forallt{\kappa} \forallt{\beta} \kappa \ \beta \rightarrow \kappa \ \beta$
which differs because the qualifier and basetype can vary independently;
see below.)\todo{Does the foregoing make sense?}

\item
  Basetype polymorphism lets the basetype vary indepentently of the
  qualifier, which might be fixed or might be polymorphic.
  An example is a nondeterministic \<choose> function, which might have type

  $\forallt{\beta} \<Set>\angles{\<Det>\ \beta} \rightarrow \<NonDet>\ \beta$

  Basetype polymorphism is rarely used.  Full type polymorphism is more
  common, even if the type is decomposed and the parts used independently,
  as in this more general signature for the \<choose> function:

  $\forallt{\kappa\ \beta} \<Set>\angles{\kappa\ \beta} \rightarrow \<NonDet> \beta$

\item
Qualifier polymorphism is common.  For example, here is the signature for
the \<length> method on strings:

$\<length>\ :\ \forallt{\kappa} . \kappa\ \<String> \rightarrow \kappa \<Int>$

\noindent
and here is that for the addition operation:

$\<plus>\ :\ \forallt{\kappa} . \kappa\ \<Int> \times \kappa\ \<Int> \rightarrow \kappa \<Int>$

\end{itemize}

\OurTypeSystem uses the standard meaning of parametric
polymorphism~\cite{Abadi:1989:FIM:77350.77373,Plotkin:1993:LPP:645891.671433}.
A polymorphic abstraction (such as a class or function) is written and
typechecked once, but can be instantiated differently at different uses,
each of which behaves as if it was written with a different non-polymorphic
type.

Informally, the \<length> above acts as if it had multiple
(overloaded) definitions

$\<NonDet>\ \<String> \rightarrow \<NonDet> \<Int>$

$\<Det>\ \<String> \rightarrow \<Det> \<Int>$

The body must typecheck at every possible instantiation, and at a use site,
the most specific applicable instantiation is chosen.
Here is an example:
(put it in a table)

\begin{Verbatim}
NonDet Int ni;
   Det Int di;

plus(ni, ni)  // has type NonDet Int; most specific applicable instantiation is NonDet Int -> NonDet Int
plus(di, di)  // has type Det Int; most specific applicable instantiation is Det Int -> Det Int
plus(ni, di)  // has type NonDet Int; most specific applicable instantiation is NonDet Int -> NonDet Int
\end{Verbatim}

We adopt the convention that  polymorphism is not instantiated in ways
that would create invalid types.  For example, the \<plus> polymorphic
function would not be instantiated at $\kappa = \<OrderNonDet>.$

\subsubsection{Polymorphism rules for collections}\label{polymorphism-up-down}

As described so far, polymorphism cannot express the collection behaviors
of \cref{sec:ond-behavior}.
Consider this typing for the \<size> function:

$\<size>\ :\ \forallt{\kappa} \kappa\ \<Collection> \rightarrow \kappa\ \<Int>$

It cannot be instantiated as

$\<size>\ :\ \<OrderNonDet>\ \<Collection> \rightarrow \<OrderNonDet> \<Int>$

because such an instantiation would include the invalid type \<OrderNonDet Int>.
A similar problem occurs for the \<first> operation that returns the first
element of a list.

\OurTypeSystem resolves this problem by introducing two variants to the
\<PolyDet> qualifier, \PolyDetUp\ and \PolyDetDown.
The $\uparrow$ operator converts \<OrderNonDet> to \<NonDet> and leaves the
other qualifiers unchanged.  The upward-pointing arrow is mnemonic of how
\<OrderNonDet> is changed:  it replaced by something higher in the type hierarchy.
The $\downarrow$ operator is analogous, but converts \<OrderNonDet> to
\<Det>, which is lower in the type hierarchy.
\Cref{fig:poly-resolutions} formalizes their behavior.

@@

%\Cref{fig-poly-resolution} presents the rule for type checking of methods annotated with \<PolyDet> or its variants.

The correct type for \<size> is

$\<size> : \forallt{\kappa} \kappa\ \<Collection> \rightarrow \down{\kappa} \<Int>$

This can be instantiated at all three type qualifiers without creating any
invalid types, and the instantiations implement the semantics of \cref{sec:ond-behavior}:

$\<size> : \<NonDet>\ \<Collection> \rightarrow \<NonDet> \<Int>$

$\<size> : \<OrderNonDet>\ \<Collection> \rightarrow \<Det> \<Int>$

$\<size> : \<Det>\ \<Collection> \rightarrow \<Det> \<Int>$

An example use of $\uparrow$ is the \<first> operation.  A simplified
version of its type is

$\<first> : \forallt{\kappa\ \beta} \kappa\ \<List>\angles{\kappa\ \beta} \rightarrow \up{\kappa}\ \beta$

whose type qualifier can be instantiated as

$\<first> : \forallt{\beta} \<NonDet>\ \<List>\angles{\<NonDet>\ \beta} \rightarrow \<NonDet>\ \beta$

$\<first> : \forallt{\beta} \<OrderNonDet>\ \<List>\angles{\<OrderNonDet>\ \beta} \rightarrow \<NonDet>\ \beta$

$\<first> : \forallt{\beta} \<Det>\ \<List>\angles{\<Det>\ \beta} \rightarrow \<Det>\ \beta$

The actual type of \<first> accounts for more possible types for the type arguments.

%\begin{figure}
%    $\infer[\rulename{poly resolution}]{\pi_r \ \beta_r\ \<m>(\overline{\pi_i \ \beta_i})\{\ \}}{\forallt{\kappa \in \{\|NonDet, OrderNonDet, Det|\}} \kappa \ \beta_r\ \<m>(\overline{K \ \beta_i})\{\ \} & \pi_r, \pi_i \in \{\|PolyDet|, \PolyDetUp, \PolyDetDown    \}}$
%    \caption{Polymorphic resolution rule}
%    \todo{I'm not sure what this is doing.  What code does it reject?}
%    \todo{This suggests that \PolyDetUp is a qualifier, rather than
%      $\uparrow$ being an operator on qualifiers.  We should clarify which
%      it is.}
%    \label{fig-poly-resolution}
%\end{figure}

@@

\begin{figure}
    $\infer[\rulename{ordernondet polyup}]
    {\up{\tau} = \|NonDet \ \beta|}
    {\tau = \|OrderNonDet \ \beta|
        & \wellformed{\tau}
    }$

@@

    $\infer[\rulename{polyup}]
    {\up{\tau} = \kappa \ \beta}
    {\tau =\kappa \ \beta
        & \kappa \in \{\|Det, NonDet|\}
        & \wellformed{\tau}
    }$

@@

    $\infer[\rulename{ordernondet polydown}]
    {\down{\tau} = \|Det \ \beta|}
    {\tau = \|OrderNonDet \ \beta|
        & \wellformed{\tau}
    }$

@@

    $\infer[\rulename{polydown}]
    {\down{\tau} = \kappa \ \beta}
    {\tau =\kappa \ \beta
        & \kappa \in \{\|Det, NonDet|\}
        & \wellformed{\tau}
    }$
\caption{Polymorphic resolution ($\uparrow$ is overloaded for types and qualifiers)}
\label{fig:poly-resolutions}
\end{figure}

@@

A simplified type for the list \<get> method for lists of strings is

$\<get> : \forallt{\kappa} \kappa\ \<List>\angles{\kappa\ \<String>} \times \kappa\ \<Int> \rightarrow \up{\kappa}\ \<String>$

(The actual typing permits the qualifier on the type argument to differ
from the type qualifier on the list, and it handles arbitrary basetypes
including collections that may themselves be order-nondeterministic.)

Instantiations are:

$\<get> : \forallt{\kappa\ \beta} \<List>\angles{\kappa\ \beta} \times \kappa\ \<Int> \rightarrow \up{\kappa}\ \beta$

$\<get> : \forallt{\kappa\ \beta} \<List>\angles{\kappa\ \beta} \times \kappa\ \<Int> \rightarrow \up{\kappa}\ \beta$

$\<get> : \forallt{\kappa\ \beta} \<List>\angles{\kappa\ \beta} \times \kappa\ \<Int> \rightarrow \up{\kappa}\ \beta$

\todo{We don't need the OrderNonDet instantiation.  The other two
  instantiations are enough.
  We don't need the following instantiation:}
\begin{Verbatim}
   // instantiation: OrderNonDet List<Det String> x _ Int -> NonDet String 
\end{Verbatim}

If either the list or the index is non-deterministic, then the result is
nondeterministic:

\begin{Verbatim}
NonDet Int ni;
   Det Int di;
NonDet List<NonDet String> nl;
OrderNonDet List<Det String> ol;
Det List<Det String> dl;

get(nl, ni) : NonDet String // instantiation: NonDet List<NonDet String> x NonDet Int -> NonDet String
get(nl, di) : NonDet String
get(ol, ni) : NonDet String
get(ol, di) : NonDet String
get(dl, ni) : NonDet String
get(dl, di) : Det String    // instantiation: NonDet List<Det String> x Det Int -> Det String
\end{Verbatim}

A similar typing does \emph{not} work for the \<set> function that sets an
element of a list, because it would permit a deterministic collection must
not be side-effected in non-deterministic ways.  Consider the following
type for

$\<set> : \forall \forallt{\tau}$

qualifier 
on the List add method: \codeid{public void add(PolyDet ArrayList<E> this, PolyDet int index, E element)}.
Suppose a client calls this method as follows:
\begin{verbatim}
@Det ArrayList<@Det String> list;
@NonDet int random;
@Det String str;
list.add(random, str);
\end{verbatim}
This will result in the instantiation \codeid{public void add(@NonDet ArrayList<E> this, @NonDet int index, E element)} and
the method invocation will type check. This is problematic because even though \<list> was declared to be \<Det>,
the type checker allows the addition of an element at a \<NonDet> location violating our determinism guarantees.
We prevent this behavior by introducing another variant of \<PolyDet>, \<PolyDet("use")> that does not affect
type instantiations. This is especially useful in preventing methods from non-deterministically modifying the state
of a deterministic receiver. To have this effect, the library method \<ArrayList.add> is annotated as:
\codeid{public void add(PolyDet ArrayList<E> this, PolyDet("use") int index, E element)}.
Since \<PolyDet("use")> has no effect on type instantiations, calling this method on a \<Det List>
will result in the method declaration being instantiated as \codeid{public void add(@Det ArrayList<E> this, @Det int index, E element)}
irrespective of the type of \<index>, thereby preventing undesired side-effects.
